---
title: Statistical Inference course project - The Effect of Vitamin C on Tooth Growth
  in Guinea Pigs
author: "Evgeniy Zabrodskiy"
date: "November, 2015"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 4, fig.width = 6)
```

```{r}
library(ggplot2)
```

## Overview
The goal of this analysis is to explore the ToothGrowth data, make assumptions and apply statistical inference methods to confirm or reject those assumptions.  
The dataset ToothGrowth contains the length of odontoblasts (teeth) in each of 60 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

## Exploratory analysis
### Basic summary of the data
ToothGrowth dataframe structure:  
```{r}
data("ToothGrowth")
str(ToothGrowth)
```
We can see that there are three variables:  
- len contains the length of the odontoblasts in microns;  
- supp is a factor variable showing the type of the supplement containing Vitamin C ("OJ" stands for "Orange Juice"", "VC" stands for "Ascrobic Acid");  
- dose stores the dose of Vitamin C in miligrams. The variable type will be converted to factor for convenience.  

Dataframe summary:  
```{r}
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
summary(ToothGrowth)
```

Two first lines of the data:  
```{r}
head(ToothGrowth, 2)
```

### Exploratory plots
The boxplot below shows the distribution of length for each *supp* per *dose*:  
```{r}
qplot(data = ToothGrowth, 
      geom = "boxplot", 
      x = dose, 
      y = len, 
      color = supp)
```

Based on this boxplot we can make different assumptions. Some of them will be described and tested below.  

## Assumptions
**Assumption 1**: Mean length of odontoblasts is different in two groups with the following properties:  
Group A: pigs which received Vitamin C with **orange juice**, dose **0.5 mg**;  
Group B: pigs which recieved Vitamin C as **ascrobic acid**, dose **0.5 mg**.  
(same dose, different supplements)  

**Assumption 2**: Mean length of odontoblasts is different in two groups with the following properites:  
Group A: pigs which received Vitamin C with **orange juice**, dose **0.5 mg**;  
Group B: pigs which received Vitamin C with **orange juice**, dose **1 mg**.  
(same supplement, different doses)  

There are more assumptions that can be made but they are all similar and are tested the same way.  

## Testing assumptions and making conclusions
Talking about lengths of odontoblasts we may assume that the lengths population is normally distributed even though we cannot clearly see that from the sample data. Based on this assumption and taking into account that the sample sizes of the groups we are going to compare is quite small (10 elements each), we should use Student's T-test to confirm or reject our assumptions. Also it is assumed that the variances of the populations of tested samples are equal since all the pigs belong to the same spicies and the study is randomized.

### Hypothesis testing for assumption 1:

$H_{0}:\mu_{oj,0.5} = \mu_{vc,0.5}$  
$H_{1}:\mu_{oj,0.5} \ne \mu_{vc,0.5}$  

```{r}
len.VC.05 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dose == 0.5]
len.OJ.05 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 0.5]
t1 <- t.test(len.VC.05, 
             len.OJ.05, 
             alternative = "two.sided", 
             paired = FALSE, 
             var.equal = TRUE)
t1
```

From the T-test output we can see that p-value = $`r t1$p.value`$ which represents the probability of $H_{0}$ being true and since it is less than $\alpha = 0.05$ for a two-sided test we may reject $H_{0}$. Also the confidence interval which does not contain zero can be used to make the same conclusion. It means that with 95% probability ($1-\alpha$) the true difference of the means of the populations will be within the borders of the interval. 

**Conclusion 1**: Assumption 1 is confirmed: mean length of odontoblasts is different for the groups of pigs which recieved 0.5 mg dose of Vitamin C in forms of orange juice and ascorbic acid.  

### Hypothesis testing for assumption 2:

$H_{0}:\mu_{oj,0.5} = \mu_{oj,1}$  
$H_{1}:\mu_{oj,0.5} \ne \mu_{oj,1}$  

```{r}
len.OJ.05 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 0.5]
len.OJ.1 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 1]
t2 <- t.test(len.OJ.05, 
             len.OJ.1, 
             alternative = "two.sided", 
             paired = FALSE, 
             var.equal = TRUE)
t2
```

From the T-test output we can see that p-value = $`r t2$p.value`$ which represents the probability of $H_{0}$ being true and since it is less than $\alpha = 0.05$ for a two-sided test we may reject $H_{0}$.  

**Conclusion 2**: Assumption 2 is confirmed: mean length of odontoblasts difference is significant for subgroups of pigs which recieved 0.5 mg and 1 mg dose of Vitamin C with orange juice.

\newpage

## Appendix - source code and additional plots

```{r, echo = TRUE, eval = FALSE}
data("ToothGrowth")
str(ToothGrowth)
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
summary(ToothGrowth)
head(ToothGrowth, 2)
```
Source code for the boxplot which is used in the analysis part of the document:  
```{r, echo = TRUE, eval = FALSE}
qplot(data = ToothGrowth, 
      geom = "boxplot", 
      x = dose, 
      y = len, 
      color = supp)
```

This boxplot is not included in the analysis part due to document length restrictions but it can be useful to view some properties of the data distribution:

```{r, echo = TRUE}
qplot(data = ToothGrowth, 
      geom = "histogram", 
      binwidth = 2, 
      x = len, 
      fill = dose, 
      facets = dose~supp)
```
